\bf Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology

### 2022-02-14 jsdse/Manuscript.tex
adding citations

Research reproducibility --the extent to which consistent results are obtained when
a scientific experiment or research workflow is repeated \citep{repdef2021}--
is a key aspect of the advancement
of science, as it constitutes a minimum standard that allows understanding research products
(e.g., methods, data, analysis, results, etc.)
to determine their reliability and generality, and eventually build up scientific
knowledge and applications based on those products
\citep{king1995replication, peng2011reproducible, powers2019open}.
In the natural sciences, rates of reproducibility are low \citep{ioannidis2005most, prinz2011believe},
which has elicited concerns about a crisis in the field \citep{baker2016reproducibility}.

In response, the scientific community has been developing new principles and standards to incentivize
cultural changes in an effort to improve long term reproducibility rates in the natural sciences
\citep{peng2015reproducibility, wilkinson2016fair}.
A standard for reproducibility that has received much attention is availability, which
we define as a property denoting that a research product can be reached (acquired, copied, analyzed,
processed and/or reused) at no financial, legal or technical cost \citep{arnold2019turing},
and without geographic, demographic, social or temporal barriers for the population \citep{fecher2014open}.

@@

To elaborate on our thesis, we designed a case study from the research field of
phylogenetics, a discipline within evolutionary biology.
We use our case study to identify barriers that have made research workflows
largely unapproachable to a general audience in the natural sciences.
Then, we propose some principles for researchers to address these barriers, and
create research workflows that are reproducible by a larger audience.
% ** describe/introduce reproducibility curriculum here
The principles proposed here can be generalized and integrated into the undergraduate
and graduate school STEM curriculum, either for courses specialized in reproducibility
or within other subject areas, as a necessary component of successful and impactful science.

@@

To explore barriers to approachability in phylogenetics, we develop a study case
that touches on three common problems within the field: standardizing
organism names in phylogenies, obtaining current phylogenetic knowledge for a group of organisms,
and summarizing this phylogenetic knowledge in a meaningful way.
To address these problems, we propose a research workflow that relies on resources from the Open
Tree of Life (OpenTree), an open source project that provides
digital availability of phylogenetic results from published, peer-reviewed research, which
is considered as vetted and state-of-the-art knowledge in the field.
OpenTree phylogenies are stored in a public database, the Phylesystem \citep{mctavish2015phylesystem},
and are downloadable as various computer-readable file types, which is key for reusability
and reproducible workflows \citep{CITATION}.
OpenTree also provides access to a single naming standard for organisms (taxonomic standard)
that is applied to the stored phylogenies \citep{rees2017automated}, which are
then used to summarize a single phylogenetic tree encompassing all life \citep{opentreeoflife2019synth}.

All of OpenTree resources
%% (the stored phylogenies, the taxonomic standard, the synthetic tree, and all
%% tools developed to process and summarize stored phylogenies),
are free of cost to any user, and are available for download and use through its Graphical User
Interface (aka, a website or application that allows users to access computer functionalities, in this case OpenTree resources, with mouse or keyboard clicks).
However, reducing as many manual steps as possible in research workflows is key
for reproducibility, as manual data manipulation scales poorly and is prone to
error \citep{bakken2019journey}.
OpenTree's resources are also programmatically available through its Application
Programming Interface services (APIs; aka computer code that implements computer functionalities,
in this case OpenTree resources, that can be used by programmers
to build more functionalities),
which provide scalability and reproducibility \citep{opentreeAPIv3}.
However, this comes at a high cost for the user, which requires considerable more
experience and literacy with computer programming to be able to successfully use APIs.
%% In recent years, OpenTree API services have been wrapped into software packages for R and Python
%% \citep{michonneau2016rotl, mctavish2021opentree}, open source, free of cost
%% programming languages that represent two of the most widely used programming languages
%% in the sciences \citep{baker2017scientific}. These packages should contribute to making
%% OpenTree resources more accessible to a wider user audience.
The R and Python programming languages are open source and free of cost, and represent
two of the most widely used programming languages in the sciences \citep{baker2017scientific}.
The rotl R package \citep{michonneau2016rotl} and the OpenTree Python module
\citep{mctavish2021opentree} are now available as wrappers for OpenTree's API services.
As such, these packages should contribute to making OpenTree resources more accessible
to a wider programming user audience.

@@

In the next section que describe the barriers to approachability that we identified
on our case study. We addressed these barriers on teaching materials that are
available at
https://mctavishlab.github.io/R\_OpenTree\_tutorials/.

@@

\section*{Identifying barriers to approchable research workflows}
\label{sec:identifying}

>>>>

Research reproducibility --the extent to which consistent results are obtained when
a scientific experiment or research workflow is repeated \citep{repdef2021}--
is a key aspect of the advancement
of science, as it constitutes a minimum standard that allows understanding research products
(e.g., methods, data, analysis, results, etc.; \cite{piwowar2013value})
to determine their reliability and generality, and eventually build up scientific
knowledge and applications based on those products
\citep{king1995replication, peng2011reproducible, powers2019open}.
In the natural sciences, rates of reproducibility are low \citep{ioannidis2005most, prinz2011believe},
which has elicited concerns about a crisis in the field \citep{baker2016reproducibility}.

In response, the scientific community has been developing new principles and standards to incentivize
cultural changes that support a long term improvement of reproducibility rates in the natural sciences
\citep{peng2015reproducibility, wilkinson2016fair, miyakawa2020no}.
A standard for reproducibility that has received much attention is availability, which
we define as a property denoting that a research product can be reached (acquired, copied, analyzed,
processed and/or reused) at no financial, legal or technical cost \citep{arnold2019turing},
and without geographic, demographic, social or temporal barriers for the population \citep{fecher2014open}.

@@

To elaborate on our thesis, we designed a case study within the research field of
phylogenetics, a discipline within evolutionary biology.
We use our case study to identify barriers that have made research workflows
largely unapproachable to a general audience in the natural sciences.
Then, we propose some principles for researchers to address these barriers, and
create research workflows that are reproducible by a larger audience.
% ** describe/introduce reproducibility curriculum here
The principles proposed here can be generalized and integrated into the undergraduate
and graduate school STEM curriculum, either for courses specialized in reproducibility
or within other subject areas, as a necessary component of successful and impactful science.

@@

To explore barriers to approachability in phylogenetics, we develop a case study
that touches on three common problems within the field: standardizing
organism names in phylogenies, obtaining current phylogenetic knowledge for a group of organisms,
and summarizing this phylogenetic knowledge in a meaningful way.
To address these problems, we propose a research workflow that relies on resources from the Open
Tree of Life (OpenTree), an open source project that provides
digital availability of phylogenetic results from published, peer-reviewed research, which
is considered as vetted and state-of-the-art knowledge in the field.
OpenTree phylogenies are stored in a public database, the Phylesystem \citep{mctavish2015phylesystem},
and are downloadable as various computer-readable file types, which is key for reusability
and reproducible workflows \citep{wilson2017good}.
OpenTree also provides access to a single naming standard for organisms (taxonomic standard)
that is applied to the stored phylogenies \citep{rees2017automated}, which are
then used to summarize a single phylogenetic tree encompassing all life \citep{opentreeoflife2019synth}.

All of OpenTree resources
%% (the stored phylogenies, the taxonomic standard, the synthetic tree, and all
%% tools developed to process and summarize stored phylogenies),
are free of cost to any user, and are available for download and use through its Graphical User
Interface (aka, a website or application that allows users to access computer functionalities, in this case OpenTree resources, with mouse or keyboard clicks).
However, reducing as many manual steps as possible in research workflows is key
for reproducibility, as manual data manipulation scales poorly and is prone to
error \citep{bakken2019journey}.
OpenTree's resources are also programmatically available through its Application
Programming Interface services (APIs; aka computer code that implements computer functionalities,
in this case OpenTree resources, that can be used by programmers
to build more functionalities),
which provide scalability and reproducibility \citep{opentreeAPIv3}.
However, this comes at a high cost for the user, which requires considerable more
computer programming experience and literacy to be able to successfully use APIs.
%% In recent years, OpenTree API services have been wrapped into software packages for R and Python
%% \citep{michonneau2016rotl, mctavish2021opentree}, open source, free of cost
%% programming languages that represent two of the most widely used programming languages
%% in the sciences \citep{baker2017scientific}. These packages should contribute to making
%% OpenTree resources more accessible to a wider user audience.
The R and Python programming languages are open source and free of cost, and represent
two of the most widely used programming languages in the sciences today \citep{baker2017scientific}.
The rotl R package \citep{michonneau2016rotl} and the OpenTree Python module
\citep{mctavish2021opentree} are now available as wrappers for OpenTree's API services.
As such, these packages should contribute to making OpenTree resources more accessible
to a wider programming user audience.

@@

In the next section que describe the barriers to approachability that we identified
on our case study. We address these barriers by developing teaching materials that are
available at
https://mctavishlab.github.io/R\_OpenTree\_tutorials/.

@@

\section*{Identifying barriers to approachable research workflows}
\label{sec:identifying}
